Effects of Reynolds Number on the Performance and Near-Wake Dynamics of a Vertical-Axis Cross-Flow Turbine

### 2015-10-10 paper.tex

\begin{document}
\todo[inline]{Rewrite abstract so it is more of a summary of what was done - 
tone down the background information a bit. (Note: some background information 
was moved to Introduction)} 
% %
\todo[inline]{Include a List of Symbols?} 
% %

% %
\listoftodos
% %

\section{Introduction}
Scaled physical models are often used in science and engineering to approximate 
real-world systems. The principle of dynamic similarity allows for
geometrically scaled systems to be dynamically identical in scale if certain
nondimensional physical parameters are matched. In the case of fluid systems,
the most important parameter is often the Reynolds number, which quantifies the
importance of viscosity on the flow. As such, hereafter we will use the word
``scale'' to refer to this dynamical scale rather than the geometric scale.

@@

Blackwell et al. investigated the effects of Reynolds number on the 
performance
of a 2 m diameter Darrieus vertical-axis cross-flow wind turbine with NACA 0012
blades \cite{Blackwell1976}. By varying wind tunnel speed, the turbine was made
to operate at approximately constant blade chord Reynolds number $Re_c$ ranging
from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine power
coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Bravo et al. observed power coefficient of a
straight-bladed VAWT to become Reynolds number independent at $Re_c \approx 4
\times 10^5$.
\todo[inline]{Add reference for Bravo et al.} 
\todo[inline]{Why did you take Polagye and Cavagnaro citation out - not 
relevant?} 
\todo[inline]{bring back the examples of studies at small scale? (too small)}

When designing or studying turbine arrays, it is common to use very small
(geometrically) scaled devices. 
% %
\todo[inline]{Get a reference for this, and a sample geometric scale.} 
% %
Sometimes these are not even models of turbines, but
wake generating objects, e.g., porous disks, that are meant to replicate the
wakes of real turbines. In this case, it is of interest to determine at what
scale one might be able to realistically study wake flows in an array, and also
to evaluate the effectiveness of a wake generator. In other words, a scale 
model wake generator may do a fine job simulating a model turbine of the same 
scale, but how well do either of them simulate a real turbine?
\todo[inline]{examples of porous disk papers, e.g. Aubrun et al?}

Previously, Bachant and Wosnik \cite{Bachant2015-JoT} showed that a cross-flow
turbine wake's unique vertical mean flow field is responsible for accelerated
wake recovery when compared with conventional axial-flow propeller-type
turbines. In this study we seek to replicate the same energy balance
considerations at multiple Reynolds numbers, to examine the implications on how
small scale, i.e. low Reynolds number experiments may be used to study flows in
turbine arrays.

Bachant and Wosnik \cite{Bachant2014} observed that maximum performance and 2-D 
near-wake profiles are affected by Reynolds number. In the present study we set 
out to analyze in greater depth the effects of scale, i.e. Reynolds number, on 
the performance of the turbine and its detailed near-wake 
characteristics. 
The goal of this study was to find a threshold scale beyond which a physical 
model study could reliably predict full-scale performance. 
%We were also looking to investigate the sensitivity as small scales, and what 
%%%the implications might be for validating
%the prediction capabilities of numerical models.
Experiments were performed that investigated Reynolds number dependence for 
scaled physical model tests of a cross-flow turbine with respect
to prediction of full-scale performance, and wake characteristics, which will be
drivers of overall performance and interaction between multiple 
turbines.

\todo[inline]{Talk about interdependence of Froude number. whatis teh Froude 
number at 1 m/s?}

@@

Hydro- or airfoil section performance is generally characterized by the 
lift-to-drag 
force ratio $L/D$  (or lift vs drag coefficient, $c_L/c_D$ ),
which is primarily a function of the foil's angle of attack $\alpha$. The 
angles of attack
for cross-flow turbine blades are a function of turbine tip speed ratio,
induction (slowing/turning of the free stream flow prior to reaching the
turbine), and turbine rotation or azimuthal angle $\theta$. Looking at static
airfoil data, we see that the static stall angle---the angle just beyond that at
which $L/D$ is maximum---increases with blade chord Reynolds number $Re_c$ 
(reference?). This
is caused by the boundary layer on the foil becoming more turbulent and more 
energetic, which enables it to advance further downstream against the adverse 
pressure gradient on the suction side and delay separation to higher angels of 
attack, as well as diminish the negative effects of a laminar separation
bubble occurring near the leading edge (reference?). For smooth airfoils, this 
transition can
cause a dramatic increase in foil performance, as seen in
Figure~\ref{fig:McMasters} at a blade chord Reynolds number on the order of
$10^5$ \cite{McMasters1980}. A review of low Reynolds number airfoil behavior is
presented in \cite{Lissaman1983} and corroborates these trends.

\todo[inline]{Laminar separation bubble - is it always there? alse need 
reference}

@@

Experiments were performed in the University of New Hampshire's tow/wave tank
in a turbine test bed specifically designed for cross-flow turbines. The 
turbine model used in this study was the UNH Reference Vertical Axis Turbine 
(RVAT), which was designed to be a generic test case for numerical
model validation, similar to the Sandia National Labs/US Department of Energy 
RM2
River Turbine, but with somewhat higher solidity, or blade chord to radius 
ratio.

@@

For each set of tows, a standard time interval was set, and each run was
analyzed to compute statistics over this interval, truncating the end slightly
to achieve an integer number of blade passages.

The experimental data and processing/plotting code are available from
\cite{Bachant2015-RVAT-Re-dep-data}.

@@

Full performance and turbine rotor drag )thrust) curves for various Reynolds 
numbers are plotted in Figure~\ref{fig:perf-curves}. 
In general, maximum $C_P$ increases with Reynolds number, from about 0.20 at 
$Re_D=0.4 \times 10^6$, to 0.26 at higher Reynolds numbers, which is due to an
increase in the foil lift-to-drag ratio. The power coefficient curves also show
a slight downward shift in the optimal tip speed ratio (peak performance) with 
increasing Reynolds 
number, from about $\lambda \approx$ 2.0 to 1.9.
This is caused by the stall delay from a more turbulent boundary later on the
blade suction side.
There is essentially no change in the shape of the
drag coefficient curves---merely an upward shift in $C_D$ with increasing $Re$.

\begin{figure}[ht]
\includegraphics[width=0.95\textwidth]{figures/perf_curves}
\caption{UNH-RVAT experiments: mean power (left) and rotor drag (right) 
coefficient curves plotted for multiple Reynolds numbers.}
\label{fig:perf-curves}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=0.95\textwidth]{figures/perf_re_dep}
\caption{UNH-RVAT experiments: mean power (left) and rotor drag (right) 
coefficients at $\lambda=1.9$ plotted versus Reynolds number.} 
\label{fig:perf-Re-dep}
\end{figure}

@@

To help understand---and possibly predict---the $Re$-sensitivity on turbine
performance, a series of static foil coefficient datasets were computed with the
XFOIL viscous panel method code \cite{Drela1989}, implemented as part of the
open source turbine design software QBlade. Simulations were run with the
default settings and zero Mach number, for the range of blade chord
Reynolds numbers encountered in the experiments.

@@

These same wake maps were measured for $Re_D = 0.4 \time 10^6$, $0.6 \times 
10^6$,
$0.8 \times 10^6$, and $1.2 \times 10^6$. Qualitatively, these wake maps look 
strikingly
similar, so they have not been plotted here. We will instead compare and
contrast the wake behavior by examining spectra and wake transport terms in the
equations that govern the downstream evolution of mean momentum and kinetic 
energy.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=0.95\textwidth]{figures/wake_spectra}
\caption{Cross-stream velocity (normalized by $U_\infty$) spectra at $z/H=0.25$ 
and $y/R=-1.0$ (or: -0.5 m in Figure~\ref{fig:wake-locations}), and  $y/R=1.5$ 
(or: +0.75 m in Figure~\ref{fig:wake-locations} (b). Dashed vertical
lines indicate $[1, 3, 6, 9]$ times the turbine rotational frequency.}
\label{fig:wake-spectra}
\end{figure}

@@

The Reynolds number dependence of the peak torque coefficient calculated from 
static foil data and cross-flow turbine kinematics was shown to be a reasonable 
indicator for predicting Reynolds number dependence of an actual cross flow 
turbine operating under dynamic conditions, including dynamic stall. The simple 
peak torque coefficient model also predicted Reynolds number independence for  
Reynolds numbers based on blade chord of $Re_c > 2 \times 10^5$.

@@

>>>>

\begin{document}
\linenumbers
\listoftodos

\todo[inline]{Rewrite abstract so it is more of a summary of what was done -
tone down the background information a bit. (Note: some background information
was moved to Introduction)}
% %
\todo[inline]{Include a List of Symbols?}

\section{Introduction}

Scaled physical models are often used in science and engineering to approximate
real-world systems. The principle of dynamic similarity allows for geometrically
scaled systems to be dynamically identical in scale if certain nondimensional
physical parameters are matched. In the case of fluid systems, the most
important nondimensional parameter is often the Reynolds number, which
quantifies the importance of inertial forces over viscous forces on the flow
physics. It should be noted that in general the geometric scale of a physical
model is not necessarily the same as its dynamical scale. As such, hereafter we
will use the word ``scale'' to refer to this dynamical scale rather than the
geometric scale.

\todo[inline]{Kinematic scaling effects? Matching TSR?}

@@

Blackwell et al. investigated the effects of Reynolds number on the
performance
of a 2 m diameter Darrieus vertical-axis cross-flow wind turbine with NACA 0012
blades \cite{Blackwell1976}. By varying wind tunnel speed, the turbine was made
to operate at approximately constant blade chord Reynolds number $Re_c$ ranging
from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine power
coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro observed high Reynolds
number sensitivity for a high solidity helical cross-flow turbine
\cite{Polagye2013b}, and Bravo et al. observed power coefficient of a
straight-bladed VAWT to become Reynolds number independent at $Re_c \approx 4
\times 10^5$.
\todo[inline]{Add reference for Bravo et al.}
\todo[inline]{Why did you take Polagye and Cavagnaro citation out - not
relevant?}
\todo[inline]{bring back the examples of studies at small scale? (too small)}

When designing or studying arrays, it is common to use very small
(geometrically) scaled devices \cite{Chamorro2011, Chamorro2011b}. It is
therefore important to realize the limitations of evaluating both the power
output of turbines and the wake recovery when Reynolds number is very far from
full scale. Sometimes models used are not turbines, but wake generating objects,
e.g., porous disks, that are meant to replicate the wakes of real turbines
\cite{Goldenberg1983}. In this case, it is of interest to determine at what
scale one might be able to realistically study wake flows in an array, and also
to evaluate the effectiveness of a wake generator. In other words, a wake
generator may do a fine job simulating a scaled turbine, but how well can it
simulate a full-scale device?

Previously, is has been observed experimentally that a cross-flow turbine wake's
unique vortical mean flow field is responsible for accelerated wake recovery
when compared with conventional axial-flow propeller-type turbines
\cite{Bachant2015-JoT}. In this study we seek to replicate the same energy
balance considerations at multiple Reynolds numbers, to examine the implications
on how small scale, i.e. low Reynolds number experiments may be used to study
flows in turbine arrays.

The Reynolds number effects on the maximum performance and 2-D near-wake
profiles of the University of New Hampshire's Reference Vertical Axis Turbine
(UNH-RVAT) were presented \cite{Bachant2014}. In this study we set out to
analyze in greater depth the effects of scale, i.e. Reynolds number, on the
performance of the turbine, and its more detailed near-wake characteristics. We
were looking to find a threshold scale beyond which a physical model study could
reliably predict full-scale performance. We were also looking to investigate the
sensitivity as small scales, and what the implications might be for validating
the prediction capabilities of numerical models.

\todo[inline]{Talk about interdependence of Froude number. What is the Froude
number at 1 m/s?}

@@

Hydro- or airfoil section performance is generally characterized by the
lift-to-drag
force ratio $L/D$  (or lift vs drag coefficient, $c_L/c_D$ ),
which is primarily a function of the foil's angle of attack $\alpha$. The
angles of attack
for cross-flow turbine blades are a function of turbine tip speed ratio,
induction (slowing/turning of the free stream flow prior to reaching the
turbine), and turbine rotation or azimuthal angle $\theta$. Looking at static
airfoil data, we see that the static stall angle---the angle just beyond that at
which $L/D$ is maximum---increases with blade chord Reynolds number $Re_c$
(reference?). This
is caused by the boundary layer on the foil becoming more turbulent and more
energetic, which enables it to advance further downstream against the adverse
pressure gradient on the suction side and delay separation to higher angels of
attack, as well as diminish the negative effects of a laminar separation
bubble occurring near the leading edge (reference?). For smooth airfoils, this
transition can
cause a dramatic increase in foil performance, as seen in
Figure~\ref{fig:McMasters} at a blade chord Reynolds number on the order of
$10^5$ \cite{McMasters1980}. A review of low Reynolds number airfoil behavior is
presented in \cite{Lissaman1983}. Note that there is a distinct lack of highly
reliable data for airfoils in this transitional regime and below. An evaluation
of the various databases relevant to cross-flow turbines is presented in
\cite{Bedon2014}.

\todo[inline]{Laminar separation bubble - is it always there? also need
reference}

@@

Experiments were performed in a turbine test bed specifically designd for cross-flow turbines. The
turbine model used in this study was the UNH Reference Vertical Axis Turbine
(RVAT), which was designed to be a generic case for numerical
model testing, similar to the Sandia National Labs/US Department of Energy RM2
River Turbine, but with higher solidity, or blade chord to radius ratio.

@@

From each set of tows, a standard time interval was set, which allowed the
turbine performance and wake to reach a quasi-periodic state. Each run was
analyzed to compute statistics over this interval, truncating the end slightly
to achieve and integer number of blade passages. Turbine shaft angular velocity
and tow carriage speed were calculated using a second order central differencing
scheme on the respective position measurements. Power and drag coefficients are
calculated as instantaneous quantities using the carriage speed as the free
stream velocity.

Wake velocity data were filtered for spurious ``spikes'' by removing datapoints
8 standard deviations or 0.9 m/s from the mean. The experimental data and code
for processing and visualization are available from
\cite{Bachant2015-RVAT-Re-dep-data}.

@@

Full power and rotor drag (a.k.a. thrust) coefficient curves for various Reynolds numbers are
plotted in Figure~\ref{fig:cp-curves} and Figure~\ref{fig:cd-curves},
respectively. There is essentially no change in the shape of the drag
coefficient curves---merely an upward shift in $C_D$ with increasing $Re$.

In general, maximum $C_P$ increases with Reynolds number, which is due to an
increase in the foil lift-to-drag ratio. The power coefficient curves also show
a slight downward shift in the optimal tip speed ratio (peak performance) with
increasing Reynolds
number, from about $\lambda \approx$ 2.0 to 1.9.
This is caused by the stall delay from a more turbulent boundary later on the
blade suction side.
There is essentially no change in the shape of the
drag coefficient curves---merely an upward shift in $C_D$ with increasing $Re$.

\begin{figure}[ht]
\centering
\includegraphics[width=0.65\textwidth]{figures/cp_curves}
\caption{Mean power coefficient curves plotted for
multiple Reynolds numbers.}
\label{fig:cp-curves}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.65\textwidth]{figures/cd_curves}
\caption{Mean rotor drag coefficient curves plotted for
multiple Reynolds numbers.}
\label{fig:cd-curves}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.9\textwidth]{figures/perf_re_dep}
\caption{UNH-RVAT measured mean power (left) and drag (right) coefficients at $\lambda=1.9$
plotted versus Reynolds number.}
\label{fig:perf-Re-dep}
\end{figure}

@@

To help understand---and possibly predict---the $Re$-sensitivity on turbine
performance, a series of static foil coefficient datasets were computed with the
XFOIL viscous panel method code \cite{Drela1989}, implemented as part of the
open source turbine design software QBlade. Simulations were run for an angle of
attack range of 0--40 degrees, in increments of 0.5 degrees. Solver parameters
used a 100 panels, a fixed speed, zero Mach number, \texttt{NCrit=9} (default
$\mathrm{e}^n$ transition criteria parameter for an average wind tunnel), and no
forced boundary layer transition. Characteristics for computed for the
approximate average blade chord Reynolds numbers encountered in the tow tank
experiments.

@@

These same wake maps were measured for $Re_D = 0.4 \times 10^6$, $0.6 \times 10^6$,
$0.8 \times 10^6$, and $1.2 \times 10^6$. Qualitatively, these look strikingly
similar, so they have not been plotted here. We will instead compare and
contrast the wake behavior by examining spectra and wake transport terms in the
equations that govern the downstream evolution of mean momentum and kinetic
energy.

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=0.85\textwidth]{figures/wake_spectra}
\caption{Cross-stream velocity (normalized by $U_\infty$) spectra at $z/H=0.25$,
$y/R=-1.0$ ($-0.5$ m in Figure~\ref{fig:wake-locations}) (a) and $y/R=1.5$ ($+0.75$ m in Figure~\ref{fig:wake-locations}) (b). Dashed vertical
lines indicate $[1, 3, 6, 9]$ times the turbine rotational frequency.}
\label{fig:wake-spectra}
\end{figure}

@@

@@

We propose a method for predicting Reynolds number dependence of a cross-flow
turbine using static airfoil characteristics and turbine blade kinematics. It is
shown that the peak torque coefficient computed this way shows similar Reynolds
number sensitivity---a better predictor than conventional quantifications of
airfoil performance, e.g., lift-to-drag ratio.
The simple peak geometric torque coefficient model predicted Reynolds number independence for
Reynolds numbers based on blade chord of $Re_c > 2 \times 10^5$.
Is it also shown from this metric
that foils with higher camber are more Reynolds number dependent but have
smaller slopes in the linear regime. The Reynolds number dependence of the peak torque coefficient calculated from
static foil data and cross-flow turbine kinematics was shown to be a reasonable
indicator for predicting Reynolds number dependence of an actual cross-flow
turbine operating under dynamic conditions, including dynamic stall.
